ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-09-24 paper/body.tex
kill me
git add .

We use both synthetic and real-world datasets from different domains, all of which are publicly available and commonly used to train and benchmark classification algorithms. We synthetically generate a dataset containing two \textit{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textit{Circles} ($n=1000$) and \textit{Moons} ($n=2500$) data. Since these data are generated by distributions of varying degrees of complexity, they allow us to assess how the generators and our proposed evaluation metrics handle this.

As for real-world data, we follow~\citet{schut2021generating} and use the \textit{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the example shown above in Figure~\ref{fig:motiv}. From the social sciences domain, we include Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on algorithmic recourse~\citep{pawelczyk2021carla}. It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether retail borrowers experience financial distress.

For the predictive modelling tasks, we use simple neural networks (\textit{MLP}) and Joint Energy Models (\textit{JEM}). For the more complex real-world datasets we also use ensembling in each case. Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}. To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset. Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target. We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for \textit{GMSC} and $n^{-}=5$ for \textit{MNIST}. Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.

\subsection{Results for Synthetic Data}

Table~\ref{tab:results-synthetic} shows the key results for the synthetic datasets separated by model (first column) and generator (second column). The numerical columns show sample averages and standard deviations of our key evaluation metrics computed across all counterfactuals. We have highlighted the best outcome for each model and metric in bold. To provide some sense of effect sizes, we have added asterisks to indicate that a given value is at least one ($*$) or two ($**$) standard deviations lower than the baseline (\textit{Wachter}).

Starting with the high-level results for our \textit{Linearly Separable} data, we find that \textit{ECCCo} produces the most faithful counterfactuals for both black-box models. This is consistent with our design since \textit{ECCCo} directly enforces faithfulness through regularization. Crucially though, \textit{ECCCo} also produces the most plausible counterfactuals for both models. This dataset is so simple that even the \textit{MLP} has learned plausible explanations of the input data. Zooming in on the granular details for the \textit{Linearly Separable} data, the results for \textit{ECCCo (no CP)} and \textit{ECCCo (no EBM)} indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM). Conformal prediction alone only leads to marginally improved faithfulness and plausibility.

The findings for the \textit{Moons} dataset are broadly in line with the findings so far: for the \textit{JEM}, \textit{ECCCo} yields substantially more faithful and plausible counterfactuals than all other generators. For the \textit{MLP}, faithfulness is maintained but counterfactuals are not plausible. This high-level pattern is broadly consistent with other more complex datasets and supportive of our narrative, so it is worth highlighting: \textit{ECCCo} consistently achieves high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility. By comparison, \textit{REVISE} yields the most plausible counterfactuals for the \textit{MLP}, but it does so at the cost of faithfulness. We also observe that the best results for \textit{ECCCo} are achieved when using both penalties. Once again though, the generative component (EBM) has a stronger impact on the positive results for the \textit{JEM}.

For the \textit{Circles} data, it appears that \textit{REVISE} performs well, but we note that it generates valid counterfactuals only half of the time (see Appendix~\ref{app:results} for a complete overview including additional common evaluation metrics). The underlying VAE with default parameters has not adequately learned the data-generating process. Of course, it is possible to improve generative performance through hyperparameter tuning but this example serves to illustrate that \textit{REVISE} depends on the quality of its surrogate. Independent of the outcome for \textit{REVISE}, however, the results do not seem to indicate that \textit{ECCCo} substantially improves faithfulness and plausibility for the \textit{Circles} data. We think this points to a limitation of our evaluation metrics rather than \textit{ECCCo} itself: computing average distances fails to account for the `wraparound' effect associated with circular data~\citep{gill2010circular}.

\import{contents/}{table-synthetic.tex}

\subsection{Results for Real-World Data}

The results for our real-world datasets are shown in Table~\ref{tab:results-real-world}. Once again the findings indicate that the plausibility attained by \textit{ECCCo} is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs. The case is very clear for \textit{MNIST}: \textit{ECCCo} consistently generates more faithful counterfactuals than other generators and plausibility gradually improves through ensembling and joint-energy modelling. Interestingly, faithfulness also gradually improves for \textit{REVISE}. This indicates that as our models improve, their generative capacity approaches that of the surrogate VAE used by \textit{REVISE}. The VAE still outperforms our classifiers in this regard, as evident from the fact that \textit{ECCCo} never quite reaches the same level of plausibility as \textit{REVISE}. With reference to Appendix~\ref{app:results} we note that the results for \textit{Schut} need to be discounted as it rarely produces valid counterfactuals for \textit{MNIST}. Relatedly, we find that \textit{ECCCo} is the only generator that consistently achieves full validity. Finally, it is worth noting that \textit{ECCCo} produces counterfactual images with the lowest average predictive uncertainty for all models.

For the tabular credit dataset (\textit{GMSC}) it is inherently challenging to use deep neural networks in order to achieve good discriminative performance~\citep{borisov2022deep,grinsztajn2022why} and generative performance~\citep{liu2022goggle}, respectively. In order to achieve high plausibility, \textit{ECCCo} effectively requires classifiers to achieve good performance for both tasks. Since this is a challenging task even for Joint Energy Models, it is not surprising to find that even though \textit{ECCCo} once again achieves state-of-the-art faithfulness, it is outperformed by \textit{REVISE} and \textit{Schut} with respect to plausibility.

\subsection{Key Takeways}

To conclude this section, we summarize our findings with reference to the opening questions. The results clearly demonstrate that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}). A related important finding is that \textit{ECCCo} yields highly plausible explanations provided that they faithfully describe model behaviour (Research Question~\ref{rq:plausibility}). \textit{ECCCo} achieves this result primarily by leveraging the model's generative property.

\import{contents/}{table-real-world.tex}

>>>>

For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks. Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}. To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset. Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.

We perform benchmarks on eight different datasets from different domains. From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit} and \textit{California Housing} (REFRENCES). All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}. Following~\citet{schut2021generating} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}. These are small image datasets but they allow us to easily visually gauge how plausible counteractuals are. (EXPLAIN WHY NO LARGER?) Finally, we also benchmark our generators on multiple synthetic datasets. Detailed descriptions and results for all datasets can be found in the appendix (POINTER). In the following we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}.

\import{contents/}{table-tabular.tex}

\import{contents/}{table-vision.tex}

\subsection{Faithfulness}

Our key finding with respect to Research Question~\ref{rq:faithfulness} is as follows:

\begin{quote}
  We find strong empirical evidence that \textit{ECCCo} consistently achieves state-of-the-art faithfulness, as it was designed to do (Research Question~\ref{rq:faithfulness}).
\end{quote}

\subsection{Balancing Objectives}

With respect to balance the objectives of faithfulness and plausibility (Research Question~\ref{rq:plausibility}) we find the following:

\begin{quote}
  The empirical evidence presented here suggests that \textit{ECCCo} consistently achieves near state-of-the-art plausibility without sacrificing faithfulness (Research Question~\ref{rq:plausibility}). 
\end{quote}

\textit{ECCCo} achieves this result primarily by leveraging the model's generative property.

### 2023-09-24 paper/tables.Rmd
git add .

```{r}
# Choices:
measures <- c(
  "unfaithfulness",
  "implausibility",
  "distance"
)
measure_names <- c(
  "Unfaithfulness ↓",
  "Implausibility ↓",
  "Cost ↓"
)
chosen_source <- "tabular"
# Order:
chosen_data <- c(
  "GMSC",
  "California Housing"
)
chosen_model <- c(
  "MLP",
  "JEM"
)

>>>>

```{r}
# Choices:
measures <- c(
  "unfaithfulness",
  "implausibility"
)
measure_names <- c(
  "Unfaithfulness ↓",
  "Implausibility ↓"
)
chosen_source <- "tabular"
# Order:
chosen_data <- c(
  "GMSC",
  "California Housing",
  "German Credit"
)
chosen_model <- c(
  "MLP",
  "JEM",
)
